---
title: "Chapter 3"
format: html
---

```{r}
library(janitor)
library(dplyr)
library(nlme)
library(ggplot2)
```

```{r}
ratpup <- read.delim(
  "data/rat_pup.dat",
  header = TRUE,
  sep = "\t"
)

# Check structure
str(ratpup)
```

```{r}
ratpup |> head()
ratpup |>
  group_by(treatment, sex) |>
  summarise(
    Obs = n(),
    Mean = mean(weight),
    StdDev = sd(weight),
    Minimum = min(weight),
    Maximum = max(weight)
  )
```

```{r}
# Calculate means for the diamond symbols
means <- aggregate(weight ~ treatment + sex, data = ratpup, FUN = mean)
```

```{r}
# Create the boxplot
ggplot(data = ratpup, aes(x = sex, y = weight)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 2) +
  # Add mean diamonds
  stat_summary(fun = mean, geom = "point", shape = 5, size = 3, stroke = 1) +
  facet_grid(~treatment) +
  scale_y_continuous(breaks = seq(4, 8, by = 1), limits = c(3, 8.5)) +
  labs(x = "sex", y = "weight") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    panel.grid = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )
```

```{r}
# Create numeric treatment variable as in SAS
ratpup$treat <- NA
ratpup$treat[ratpup$treatment == "High"] <- 1
ratpup$treat[ratpup$treatment == "Low"] <- 2
ratpup$treat[ratpup$treatment == "Control"] <- 3

# Encoding
ratpup$sex <- factor(ifelse(ratpup$sex == "Male", 1, 0))
```

```{r}
# Fit initial linear mixed model
model3.1.fit <- lme(
  weight ~ treatment + sex + lsize + treatment:sex,
  random = ~ 1 | litter,
  data = ratpup,
  method = "REML"
)
```

$$
\begin{aligned}
weight_{ij} &=
(\beta_0 + b_j)
+ \beta_1\,treatment_{ij}
+ \beta_2\,sex_{ij}
+ \beta_3\,litsize_{ij}
+ \beta_4\,(treatment_{ij} \times sex_{ij})
+ \varepsilon_{ij} \\
\\
b_j &\sim \mathcal{N}(0,\sigma_b^2), \quad j = 1,\dots,J \\
\varepsilon_{ij} &\sim \mathcal{N}(0,\sigma^2)
\end{aligned}
$$

```{r}
# ratpup$treatment <- relevel(ratpup$treatment, ref = "High")
summary(model3.1.fit)
```

F-tests for the fixed effects in the model.

```{r}
anova(model3.1.fit)
```

The `anova()` function performs a series of `Type I F`-tests (or sequential) for the fixed effects in the model, each of which is conditional on the preceding terms in the model specification.  

For example, the `Type I F`-test for `SEX1` is conditional on the `TREATMENT` effects, but the `Type I F`-test for `TREATMENT` is not conditional on the `SEX1` effect.

The random.effects() function can be used to display the EBLUPs for the random
litter effects:

```{r}
# Display the random effects (EBLUPs) from the model.
random.effects(model3.1.fit)
```